Fail Faster

### 2025-03-20 paper/paper.tex
nits

For \camlname, our test cases consist of generators for boolean lists, binary search trees (BST), and simply
typed lambda calculus (STLC) terms. We implement generators for these benchmarks
using a variety of \textit{strategies}, which vary in structure and
sophistication. These include type-derived generators for BSTs and STLC terms,
using the approach outlined in \secref{subsection:type-derived}; two custom BST
generators, one that repeatedly inserts values into an initially empty tree and
one that materializes the tree in a single pass by generating keys, values, and
subtrees at each step; a generator of boolean lists that operates similarly to
the single-pass BST strategy; and a custom, correct-by-construction generator of
STLC terms, whose behavior is the most complex. 
For evaluating \scalaname, we implement a subset of the afformentioned 
strategies---only the boolean list generator, single-pass BST generator, and well-typed STLC term generator.

The runtime of the \camlname generators over a range of sizes is shown in
\figref{fig:time}. Staging provides only a negligible speedup for type-derived
and insertion-based BST strategies, whose overhead is dominated by random
sampling. In this case, performance improvement is primarily attributable the
\csplitmix \rand.  In contrast, STLC strategies benefit more from staging, as
they involve fewer sampling operations but rely heavily on generator
combinators. The same holds for boolean lists, as sampling booleans is
inexpensive relative to integers. Our single-pass BST strategy falls somewhere
in between: it both samples frequently and makes extensive use of generator
combinators. \tr{Goal is to replace this paragraph with the ref-counting
chart---I think it will be WAY less boring. But I need to do 3.7 first.}

@@

%\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
%\jwc{
%  Baseline generators to test speed in both languages:
%  \begin{itemize}
%    \item Single int
%    \item Pair of ints, constrained
%    \item List of ints without bind (use map): lots of sampling, minimal binds.
%    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
%  \end{itemize}
%}
\subsection{Benchmarking Bug-Finding Ability}
To answer \textbf{RQ2}, we evaluate the bug-finding ability of our BST and STLC
case studies using Etna, a platform for assessing and comparing PBT techniques. 
Etna allows users to measure the effectiveness of different generator
implementations by injecting bugs into the system under test and recording the
time taken for a relevant property to fail in response. For example, in a tree
with an ordering property, a bug injection might consist of inserting a value
into the tree that violates that ordering. Each case study includes a diverse
set of \textit{tasks}, where a task consists of a specific bug-property pair
designed to test different aspects of the system (e.g., BST includes tasks that
test insertion, deletion, and union operations on binary search trees). In our
case, BST has 43 tasks, and STLC has 18.

Strategies for BST and STLC are unchanged from \secref{subsection:benchmarking}.
To ensure a fair comparison, we initialize each strategy with the same random
seed, ensuring that they produce identical sequences of values. We run each
strategy on all tasks using a 60-second timeout.  If a bug is not found within
this limit, the task is considered unsolved.  Tasks where all strategies fail to
find the bug are excluded from our dataset, as they provide no basis for
comparison.  Similarly, we exclude tasks where the BQ strategy completes in
under 5ms, as such negligible runtimes do not yield meaningful insights into
relative performance. These filters ensure that reported speedups are based on
cases where optimization has a measurable impact.

It is worth noting that the time-to-failure of a given strategy on a given seed
is not necessarily representative of the strategy's average time-to-failure over
a large number of trials. We normalize by computing the relative performance, or
the ``speedup''. This works in most cases, but it is not perfect.  For example,
it is theoretically possible to choose a seed such that the \textit{first} value
produced by a generator discovers the bug, which would elide any differential
speedup (and likewise if both generators fail to find the bug).  To account for
these cases, we repeat the above process over 10 seeds.\footnote{Although the
variance \textit{between} seeds in Etna can be vast, timing results are
replicable for a \textit{given} seed. Across 1,000 trials using the same seed, 
the observed variance in time-to-failure was less than a nanosecond.}

The average individual-task speedups for each strategy and benchmark are shown in 
\figref{fig:etna}, with trends in bug-finding ability closely mirroring those in performance 
from \secref{subsection:benchmarking}. The distribution of speedups in \figref{fig:swarm} reveals 
substantial variability, with most tasks clustering near the median and a long tail of outliers 
achieving much larger gains. STLC shows a more bimodal distribution of speedups
than the other benchmarks, which we attribute to the heterogenous difficulty of its tasks:  
some tasks regularly hit the 60-second timeout, while others finish in a fraction of a second. We find that tasks of intermediate difficulty 
tend to achieve larger speedups, whereas very fast or slow tasks cluster near the mean.
Overall, 
these results give strong evidence that our technique consistently benefits bug-finding ability, sometimes drastically.
\begin{figure}[h]
  \includegraphics[scale=0.333]{speedup.png}
  \caption{Geometric average of all speedups for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board.}
  \label{fig:etna}
\end{figure}

>>>>

For \camlname, our test cases consist of generators for boolean lists, binary search trees (BST), and simply
typed lambda calculus (STLC) terms. We implement generators for these benchmarks
using a variety of \textit{strategies}, which vary in structure and
sophistication. These include type-derived generators for BSTs and STLC terms,
using the approach outlined in \secref{subsection:type-derived}; two custom BST
generators, one that repeatedly inserts values into an initially empty tree and
one that materializes the tree in a single pass by generating keys, values, and
subtrees at each step; a generator of boolean lists that operates similarly to
the single-pass BST strategy; and a custom, correct-by-construction generator of
STLC terms, whose behavior is the most complex. 
For evaluating \scalaname, we implement a subset of the aformentioned 
strategies---only the boolean list generator, single-pass BST generator, and well-typed STLC term generator.

The runtime of the \camlname generators over a range of sizes is shown in
\figref{fig:time}. Staging provides only a negligible speedup for type-derived
and insertion-based BST strategies, whose overhead is dominated by random
sampling. In this case, performance improvement is primarily attributable the
\csplitmix \rand.  In contrast, the STLC strategies benefit more from staging, as
they involve fewer sampling operations but rely heavily on generator
combinators. The same holds for boolean lists, as sampling booleans is
inexpensive relative to integers. Our single-pass BST strategy falls somewhere
in between: it both samples frequently and makes extensive use of generator
combinators. \tr{Goal is to replace this paragraph with the ref-counting
chart---I think it will be WAY less boring. But I need to do 3.7 first.}

@@

%\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
%\jwc{
%  Baseline generators to test speed in both languages:
%  \begin{itemize}
%    \item Single int
%    \item Pair of ints, constrained
%    \item List of ints without bind (use map): lots of sampling, minimal binds.
%    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
%  \end{itemize}
%}
\subsection{Benchmarking Bug-Finding Ability}
To answer \textbf{RQ2}, we evaluate the bug-finding ability of our BST and STLC
case studies using Etna, a platform for assessing and comparing PBT techniques. 
Etna allows users to measure the effectiveness of different generator
implementations by injecting bugs into the system under test and recording the
time taken for a relevant property to fail in response. For example, in a tree
with an ordering property, a bug injection might consist of inserting a value
into the tree that violates that ordering. Each case study includes a diverse
set of \textit{tasks}, where a task consists of a specific bug-property pair
designed to test different aspects of the system (e.g., BST includes tasks that
test insertion, deletion, and union operations on binary search trees). In particular, BST has 43 tasks, and STLC has 18.

Strategies for BST and STLC are unchanged from \secref{subsection:benchmarking}.
To ensure a fair comparison, we initialize each strategy with the same random
seed, ensuring that they produce identical sequences of values. We run each
strategy on all tasks using a 60-second timeout.  If a bug is not found within
this limit, the task is considered unsolved.  Tasks where all strategies fail to
find the bug are excluded from our dataset, as they provide no basis for
comparison.  Similarly, we exclude tasks where the BQ strategy completes in
under 5ms, as such negligible runtimes do not yield meaningful insights into
relative performance. \tr{We probably want to discuss how many tasks get filtered out...} These filters ensure that reported speedups are based on
cases where optimization has a measurable impact.

It is worth noting that the time-to-failure of a given strategy on a given seed
is not necessarily representative of the strategy's average time-to-failure over
a large number of trials. We normalize by computing the relative performance, or
the ``speedup''. This works in most cases, but it is not perfect.  For example,
it is theoretically possible to choose a seed such that the \textit{first} value
produced by a generator discovers the bug, which would elide any differential
speedup (and likewise if both generators fail to find the bug).  To account for
these cases, we repeat the above process over 10 \tr{(soon to be 30)} seeds.\footnote{Although the
variance \textit{between} seeds in Etna can be vast, timing results are
replicable for a \textit{given} seed. Across 1,000 trials using the same seed, 
the observed variance in time-to-failure was less than a nanosecond.}

The average individual-task speedups for each strategy and benchmark are shown in 
\figref{fig:etna}, with trends in bug-finding ability closely mirroring those in performance 
from \secref{subsection:benchmarking}. The distribution of speedups in \figref{fig:swarm} reveals 
substantial variability, with most tasks clustering near the median and a long tail of outliers 
achieving much larger gains. STLC shows a more bimodal distribution of speedups
than the other benchmarks, which we attribute to the heterogenous difficulty of its tasks:  
some tasks regularly hit the 60-second timeout, while others finish in a fraction of a second. We find that tasks of intermediate difficulty 
tend to achieve larger speedups, whereas very fast or slow tasks cluster near the mean. \tr{To be honest, we don't know EXACTLY why STLC exhibits this bimodal distribution. Maybe it's better to leave this out, and just say ``look, there's high variance, but it's actually a good thing---sometimes you get more than you expect, and you almost never get less than you expect"?}
Overall, 
these results give strong evidence that our technique consistently benefits bug-finding ability, sometimes drastically.
\begin{figure}[h]
  \includegraphics[scale=0.333]{speedup.png}
  \caption{Geometric average of all speedups for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board.}
  \label{fig:etna}
\end{figure}
